On the Minimum Core Mass for Giant Planet Formation

### 2013-07-30 tex/draft/draft3.tex
section 4 responses to Ana's comments

\begin{figure}[tb]
\centering
%AY: figure should probably use R_B for consistency with text.
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/PTm_profiles_v2.pdf}
%\vspace{-0.5in}
\caption{Radial profiles of atmospheric pressure, temperature and enclosed mass (including the core) for a $5 M_{\oplus}$ core at $a=60$ AU.   Solid, dotted and dashed lines correspond to solutions with total mass (core and atmosphere) of $5.12 M_{\oplus}$, $5.85 M_{\oplus}$ and $9.22 M_{\oplus}$, respectively.  Circles and triangles mark the locations of the Bondi radii and of the radiative-convective boundaries, respectively.  The radial profiles extend from the core to the Hill radius.} 
\label{fig:profiles}
\end{figure}

@@

At later stages of evolution, the numerical model gives a flat luminosity with increasing mass and also time (not shown).  By contrast the analytic model gives a luminosity that continues to drop with increasing mass.  The difference is due to the neglect of self-gravity in the analytic model.  With $L\cb \propto M\cb T\cb^4/ (\kappa\cb  P\cb)$, accounting for the higher enclosed mass gives a somewhat higher luminosity in the self-gravitating model.  However the main effect is the higher temperature in the self-gravitating models, which have the flatter temperature profiles as described above.

@@

\Fig{fig:LtvsMopacity} shows the effect of decreasing the atmospheric dust abundance on the atmospheric evolution.  The opacity normalization in \Eq{eq:opacitylaw} is reduced by factors of 10 and 100.  Lower opacities result in higher luminosities and faster evolution, a well-established result that we confirm for our model.  Atmospheric dust opacities are difficult to robustly predict. Ablation of infalling solids is a dust source, while settling through the radiative zone is a sink.  Grain growth both reduces dust opacities per unit mass and favors settling.  Our scenario of negligible ongoing particle accretion tends to favor low dust opacities.   To be conservative, however, our reference case considers full Solar abundances, but faster evolution is clearly possible.

@@

\Fig{fig:growthtime} plots the evolution of the atmospheric growth timescale, $M_{\rm atm}/\dot{M}$,  around a $5 M_{\oplus}$ core at several locations in the disk. These growth times are short both early -- when the radiative zone is relatively transparent -- and late -- when self-gravity accelerates growth.  Thus the time required to accrete an atmosphere is dominated by intermediate masses, $M_{\rm atm} \sim 1-3 M_{\oplus}$ for the chosen parameters.  The accelerated growth at higher masses is also evident in \Fig{fig:Ltplot} (bottom panel) as the flattening of the $t$ vs.\ $M_{\rm atm}$ curve.

Generically for the models we considered, growth times peaked for $M_{\rm{atm}} < M_{\rm c}$, i.e.\ before the crossover mass.  Thus cumulative accretion timescales can be well approximated by only considering the evolution prior to the crossover mass.  This result is fortuitous, as we next show how our model assumptions can break down during the later stages of atmospheric growth.

@@

 \Fig{fig:coolingterms} shows that the neglected luminosity is indeed negligible during the early stages of evolution.  However, $L_{\rm negl}$ exceeds the model luminosity, $L$, for $M_{\rm atm} > 3 M_\oplus$, meaning that our model is becoming inaccurate.  Nevertheless, our cooling model gives both conservative and reasonably reliable estimates of core accretion timescales.  Our results are conservative -- in the sense of overestimating growth times -- because the neglected luminosity would give more rapid evolution.  And our results are reasonably accurate because the bulk of growth time is spent at low masses, when $L_{\rm negl}$ is small and safely neglected.

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cooling_a10_Mc5_rcb_paper.pdf}
%\vspace{-0.5in}
\caption{Luminosity evolution around a $5 M_\oplus$ core at 10 AU.  In addition to the total model luminosity $L$, the contributing terms in the cooling model of \Eq{eq:coolingglobal}; $-\dot{E}$, $e_{\rm M} \dot{M}$ and $-P \p V/ \p t |_M$; are shown.  The $L_{\rm negl}$ curve gives the neglected luminosity that would have been generated in the radiative zone, but is not included in our two-layer cooling model.}
\label{fig:coolingterms}
\end{figure}

@@

\subsubsection{Disk Temperature and Pressure}
\label{sec:TPeffects}

>>>>

\begin{figure}[tb]
\centering
%AY: figure should probably use R_B for consistency with text.
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/PTm_profiles_v2.pdf}
%\vspace{-0.5in}
\caption{Radial profiles of atmospheric pressure, temperature and enclosed mass (including the core) for a $5 M_{\oplus}$ core at $a=60$ AU.   Solid, dotted and dashed lines correspond to solutions with total mass (core and atmosphere) of $5.12 M_{\oplus}$, $5.85 M_{\oplus}$ and $9.22 M_{\oplus}$, respectively.  (See text for a description of evolution to yet higher masses.)   Circles and triangles mark the locations of the Bondi radii and of the radiative-convective boundaries, respectively.  The radial profiles extend from the core to the Hill radius.} 
\label{fig:profiles}
\end{figure}

@@

At later stages of evolution, the numerical model gives a flat luminosity with increasing mass and also time (not shown).  By contrast the analytic model gives a luminosity that continues to drop with increasing mass.  The difference is due to the neglect of self-gravity in the analytic model.  From \Eq{eq:Lcb}, $L\cb \propto M\cb T\cb^4/ (\kappa\cb  P\cb)$.  Thus accounting for the higher enclosed mass gives a somewhat higher luminosity in the self-gravitating model.  However the main effect is the higher temperature in the self-gravitating models.  The flatter temperature profiles with radius, as described above, explain why the self-gravitating models are hotter.

@@

\Fig{fig:LtvsMopacity} shows the effect of decreasing the atmospheric dust abundance on the atmospheric evolution.  The opacity normalization in \Eq{eq:opacitylaw} is reduced by factors of 10 and 100.  Lower opacities result in higher luminosities and faster evolution, a well-established result \citep{HubBod05} that we confirm for our model.  Atmospheric dust opacities are difficult to robustly predict. Ablation of infalling solids is a dust source, while settling through the radiative zone is a sink.  Grain growth both reduces dust opacities per unit mass and favors settling.  Our scenario of negligible ongoing particle accretion tends to favor low dust opacities.   To be conservative, however, our reference case considers full Solar abundances, but faster evolution is clearly possible.

@@

\Fig{fig:growthtime} plots the evolution of the atmospheric growth timescale, $M_{\rm atm}/\dot{M}$,  around a $5 M_{\oplus}$ core at several locations in the disk.  This instantaneous growth time shows clearly that the atmosphere spends the bulk of its time growing though intermediate atmospheric masses, $\sim 1 -3$ $M_\oplus$ in this case.  Growth times are short both early -- when the radiative zone is relatively transparent -- and late -- when self-gravity accelerates growth.  The accelerated growth at higher masses is also evident in \Figs{fig:Ltplot}{fig:LtvsMopacity} (bottom panels) as the flattening of the $t$ vs.\ $M_{\rm atm}$ curve.

The fact that growth times peak before the crossover mass, when $M_{\rm{atm}} < M_{\rm c}$, is very helpful for understanding atmospheric growth.  The estimated timescale to runaway growth can be well approximated by only considering the time it takes to grow to the crossover mass, or even to a somewhat lower mass.  This insensitivity to the upper mass threshold is characteristic of accelerating growth, and is fortuitious.  We next show how our model assumptions can break down during the later stages of atmospheric growth.

@@

 \Fig{fig:coolingterms} shows that the neglected luminosity is indeed negligible during the early stages of evolution.  However, $L_{\rm negl}$ exceeds the model luminosity, $L$, at high masses, $M_{\rm atm} > 3 M_\oplus$ in this case, meaning that our model is becoming inaccurate.  Nevertheless, our cooling model gives both conservative and reasonably reliable estimates of core accretion timescales.  Our results are conservative -- in the sense of overestimating growth times -- because the neglected luminosity would give more rapid evolution.  Moreover our results are reasonably accurate because the bulk of growth time is spent at low masses, when $L_{\rm negl}$ is small and safely neglected.

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cooling_a10_Mc5_rcb_paper.pdf}
%\vspace{-0.5in}
%\caption{Luminosity evolution around a $5 M_\oplus$ core at 10 AU.  In addition to the total model luminosity $L$, the contributing terms in the cooling model of \Eq{eq:coolingglobal}; $-\dot{E}$, $e_{\rm M} \dot{M}$ and $-P \p V/ \p t |_M$; are shown.  The $L_{\rm negl}$ curve gives the neglected luminosity that would have been generated in the radiative zone, but is not included in our two-layer cooling model.}
\label{fig:coolingterms}
\end{figure}

@@

\subsection{Disk Temperature and Pressure}
\label{sec:TPeffects}
